---
title: "Capstone Part 2: Choose your own"
author: "Jeremy"
date: "25/08/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# **Preface**

The following machine learning (ML) project is part of the Harvardx Data Science: Capstone course, where learners are given an opportunity to showcase ML within R-programming of one’s choosing. 

# **Introduction**
As computer and technological systems improve, many different sectors use machine learning to do what cannot be physically done by a human being within short time spans. Within the field of abnormal psychology, there are also many opportunities to apply machine learning. With the increase in detecting many facets of the human brain, machine learning can process the data captured to identify different psychopathologic disorders. Thus, ML can help in prioritising patients in need of further testing, and if so, lead to timely interventions.   


In this analysis, I will be using a UCI Parkinsons data set as the basis. This data set consists of 23 speech attributes from 195 voice recordings. From this set, I will be dividing the data set into two, and once I trained the algorithm on the training set, I will apply it to the test set to get the best accuracy. As the result is a categorical result, the two main methods used is K-nearest neighbours (KNN) and random forest. The results will show the method used, the accuracy obtained from the algorithm, and the F1 score. 

# **Methods and analysis**

Here, the packages used are loaded into the script. The dataset is downloaded directly from the database.

https://archive.ics.uci.edu/ml/datasets/Parkinsons

# 1. Packages and data Load in

```{r, message = F}
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")
if(!require(rpart)) install.packages("rpart", repos = "http://cran.us.r-project.org")
if(!require(randomForest)) install.packages("randomForest", repos = "http://cran.us.r-project.org")
if(!require(caretEnsemble)) install.packages("caretEnsemble", repos = "http://cran.us.r-project.org")
if(!require(kableExtra)) install.packages("kableExtra", repos = "http://cran.us.r-project.org")

dat <- read_csv("https://archive.ics.uci.edu/ml/machine-learning-databases/parkinsons/parkinsons.data")
```

# 2. Dividing into train and test sets 

Here, the dataset is split into 2 partitions, by a 50/50 split. As the data set is relatively small, the 50:50 split ensures that there is equal amount for the algorithm to work on, before it can be tested on the test set. Also, a further analysis of the data set shows that there is no even balance between healthy and parkison's patients. Thus, the prevalence narrows how the data set should be split. Later in the analysis, the randomforest method will elaborate on k-fold methods. 

```{r, message = F}
dat <- as.data.frame(dat)
set.seed(1, sample.kind="Rounding")
index <- createDataPartition(dat$status, times = 1, p = 0.5, list = F)

test_dat <- dat[index,]
train_dat <- dat[-index,]
```

# 3. Preparing the training dataset

These terms were given by the UCI source on measures used for detecting Parkinson's.

name - ASCII subject name and recording number
MDVP:Fo(Hz) - Average vocal fundamental frequency
MDVP:Fhi(Hz) - Maximum vocal fundamental frequency
MDVP:Flo(Hz) - Minimum vocal fundamental frequency
MDVP:Jitter(%),MDVP:Jitter(Abs),MDVP:RAP,MDVP:PPQ,Jitter:DDP - Several measures of variation in fundamental frequency
MDVP:Shimmer,MDVP:Shimmer(dB),Shimmer:APQ3,Shimmer:APQ5,MDVP:APQ,Shimmer:DDA - Several measures of variation in amplitude
NHR,HNR - Two measures of ratio of noise to tonal components in the voice
status - Health status of the subject (one) - Parkinson's, (zero) - healthy
RPDE,D2 - Two nonlinear dynamical complexity measures
DFA - Signal fractal scaling exponent
spread1,spread2,PPE - Three nonlinear measures of fundamental frequency variation

However, the column names will be renamed for better processing and understanding. 

Subject - ASCII subject name and recording number
Avg.vocal.freq - Average vocal fundamental frequency
Max - Maximum vocal fundamental frequency
Min.vocal.freq - Minimum vocal fundamental frequency
Jitter.perc,Jitter.Abs,RAP,PPQ,DDP - Several measures of variation in fundamental frequency
Shimmer,Shimmer.dB,Shimmer.APQ3,Shimmer.APQ5,APQ,DDA - Several measures of variation in amplitude
NHR,HNR - Two measures of ratio of noise to tonal components in the voice
status - Health status of the subject (one) - Parkinson's, (zero) - healthy
RPDE,D2 - Two nonlinear dynamical complexity measures
DFA - Signal fractal scaling exponent
spread1,spread2,PPE - Three nonlinear measures of fundamental frequency variation

```{r}
#renaming columns
names(train_dat) <- c("Subject", 
                      "Avg.vocal.freq",
                      "Max.vocal.freq",
                      "Min.vocal.freq",
                      "Jitter.perc",
                      "Jitter.Abs",
                      "RAP",
                      "PPQ",
                      "DDP",
                      "Shimmer",
                      "Shimmer.dB",
                      "Shimmer.APQ3",
                      "Shimmer.APQ5",
                      "APQ",
                      "Shimmer.DDA",
                      "NHR",
                      "HNR",
                      "status",
                      "RPDE",
                      "D2",
                      "DFA",
                      "Spread1",
                      "Spread2",
                      "PPE")

train_dat <- as.data.frame(train_dat)
train_dat_numeric <- train_dat %>% select(-"Subject", -"status")

colnames <- colnames(train_dat_numeric)

#preparing the test set:
names(test_dat) <- c("Subject", 
                      "Avg.vocal.freq",
                      "Max.vocal.freq",
                      "Min.vocal.freq",
                      "Jitter.perc",
                      "Jitter.Abs",
                      "RAP",
                      "PPQ",
                      "DDP",
                      "Shimmer",
                      "Shimmer.dB",
                      "Shimmer.APQ3",
                      "Shimmer.APQ5",
                      "APQ",
                      "Shimmer.DDA",
                      "NHR",
                      "HNR",
                      "status",
                      "RPDE",
                      "D2",
                      "DFA",
                      "Spread1",
                      "Spread2",
                      "PPE")

test_dat <- as.data.frame(test_dat)
```

## 3.1 Exploring the training dataset

In this section we will see how the scores are distributed. Initially, the histogram does not give much information other than the spread of the predictors. However, when visualised by comparing the status (healthy vs parkinson's), clear distinctions in the scores form. 

```{r}
#data set
head(train_dat)

#visualising the distribution of scores
for (i in 1:22) {
    hist(train_dat_numeric[,i], probability=TRUE,
         main = colnames[i],
         xlab = NULL)
}


#visualising if values change according to status
#part 1:
a <- train_dat %>% 
  select(Avg.vocal.freq:Shimmer.APQ3, status) %>% 
  mutate(status = as.factor(status)) %>% 
  gather(comp, value, -status) %>% 
  ggplot(aes(comp,value, fill = status))+
  geom_boxplot()+
  facet_wrap(~comp, scales = "free")

#part 2:
b <- train_dat %>% 
  select(Shimmer.APQ3:PPE) %>% 
  mutate(status = as.factor(status)) %>% 
  gather(comp, value, -status) %>% 
  ggplot(aes(comp,value, fill = status))+
  geom_boxplot()+
  facet_wrap(~comp, scales = "free")

a;b

```

From here, the following components have visually distinct differences between healthy and Parkinson's recordings:
- Jitter.Abs, Jitter.perc, PPQ, Shimmer, Shimmer.dB, APQ, DFA, Shimmer.APQ3, Shimmer.APQ5, Shimmer.DDA, Spread1

# **Analysis: Fitting the algorithms**

# 4. Fitting models: KNN

K-nearest neighbours, or KNN, estimates the average of the observations used (the neighbourhood). The flexibility in the method comes from the k parameter, which we will see fine tuned in this section. Here, *all* predictors are used for the analysis. 

```{r}
#prep work
train_dat1 <- train_dat %>% 
  select(-Subject) %>% 
  mutate(status = as.factor(status))
test_dat1 <- test_dat %>% select(-Subject) %>% 
  mutate(status = as.factor(status))

ks <- seq(1, 96, 2)

accuracy <- map_df(ks, function(k){
  fit <- knn3(status ~ ., data = train_dat1, k = k)
  y_hat <- predict(fit, train_dat1, type="class")
  cm_train <- confusionMatrix(y_hat, train_dat1$status)
  train_error <- cm_train$overall["Accuracy"]

  y_hat <- predict(fit, test_dat1,type="class")
  cm_test <- confusionMatrix(y_hat, test_dat1$status)
  test_error <- cm_test$overall["Accuracy"]
  
  tibble(train = train_error, test = test_error)
})

accuracy
which.max(accuracy$test) #when k = 5, accuracy highest

#using k=5
fit.knn <- knn3(status ~ ., data = train_dat1, k = 5)
y_hat_knn <- predict(fit.knn, test_dat1, type="class")
cm_test <- confusionMatrix(y_hat_knn, test_dat1$status)
cm_test$overall["Accuracy"] #0.82653

#checking F1 scores
F_meas(y_hat_knn,test_dat1$status) #0.58537

#prevalence in test_dat
mean(1-test_dat$status)

acc_results <- data_frame(Method = "Tuned KNN (k=5)", #column 1: method
                           Accuracy = cm_test$overall["Accuracy"],
                          F1.score  = F_meas(y_hat_knn,test_dat1$status))

acc_results

#prevalence
mean(train_dat$status == 1)
```

## 4.1 Tuning the F1 score

As in the last section, we see that the F1 score is quite low. The F1 score refers to the harmonic average between both specificity and sensitivity.  Calculating the F1 score here is important as the prevalence of the status is not an exact 50/50 split (77.3% of the set has detected parkinsons). A low F1 score could mean that the algorithm have poor precision and poor recall. 

Here, we find that the F1 score can improve.

```{r}
ks <- seq(1, 50, 2)
Fvalue <- map_df(ks, function(k){
  fit.knn <- knn3(status ~ ., data = train_dat1, k = k)
  y_hat_knn <- predict(fit.knn, test_dat1, type="class")
  cm_train <- confusionMatrix(y_hat_knn, test_dat1$status)
  Fvalue <- F_meas(y_hat_knn,test_dat1$status)
  
  tibble(Fvalue)
})
Fvalue
ks[which.max(Fvalue$Fvalue)] #when k = 1, F1 score is 0.73077
```

## 4.2 Using components with clear distinctions

Here, KNN is used with tuning of the k value for *selected* predictors. 

As seen in section 3:
The following components have visually distinct differences between healthy and parkinson-recordings:
- Jitter.Abs, Jitter.perc, PPQ, Shimmer, Shimmer.dB, APQ, DFA, Shimmer.APQ3, Shimmer.APQ5, Shimmer.DDA, Spread1

```{r}
#selecting only components with clear distinctions
train_dat_special <- train_dat1 %>% select(Jitter.Abs, Jitter.perc, PPQ, Shimmer, Shimmer.dB, Shimmer.APQ3, Shimmer.APQ5, Shimmer.DDA, APQ, DFA, Spread1, status)
test_dat_special <- test_dat1 %>% select(Jitter.Abs, Jitter.perc, PPQ, Shimmer, Shimmer.dB, Shimmer.APQ3, Shimmer.APQ5, Shimmer.DDA, APQ, DFA, Spread1, status)

#tuning knn
ks <- seq(1, 96,2)
accuracy <- map_df(ks, function(k){
  fit <- knn3(status ~ ., data = train_dat_special, k = k)
  y_hat <- predict(fit, test_dat1, type="class")
  cm_train <- confusionMatrix(y_hat, test_dat_special$status)
  test_error <- cm_train$overall["Accuracy"]
  tibble(test = test_error)
})
accuracy
ks[which.max(accuracy$test)] #accuracy is highest at k = 9

#using k = 9
fit.knn.s <- knn3(status ~ ., data = train_dat_special, k = 9)
y_hat_knn_s <- predict(fit.knn.s, test_dat1, type="class")
cm_test_knn_s <- confusionMatrix(y_hat_knn_s, test_dat_special$status)
cm_test_knn_s$overall["Accuracy"]
#accuracy at 0.8776

#tuning F1 score
F.value_s <- map_df(ks, function(k){
  fit <- knn3(status ~ ., data = train_dat_special, k = k)
  y_hat <- predict(fit, test_dat, type="class")
  cm_train <- confusionMatrix(y_hat, test_dat_special$status)
  Fvalue.s <- F_meas(y_hat,test_dat1$status)
  
  tibble(Fvalue.s)
})
F.value_s[5,]

ks[which.max(F.value_s$Fvalue.s)] #when k = 9, F1 score is 0.73913 (highest)

#inputting the results
acc_results1 <- bind_rows(acc_results, 
                          data.frame (Method = "Selected predictors, Tuned KNN (k=9)", 
                           Accuracy = cm_test_knn_s$overall["Accuracy"],
                          F1.score  = F_meas(y_hat_knn_s,test_dat1$status)))


acc_results1
```
Therefore, using components that are visually distinct can offer better accuracy and F1 scores to the results. 

# 5. Fitting models: Random forest

Random forest (RF) is a classification method that is a step up from classification trees, which generates a single decision tree from the data. To address the shortcomings of decision trees, the rf method averages multiple decision trees. The flexibility in the method comes from the mtry parameter, which is the number of variables randomly sampled as candidates at each split. 

Here, random forest is used for *selected* predictors. 

```{r}
#random forest without tuning
fit.rf <- randomForest(status~., data = train_dat_special)
fit.rf;plot(fit.rf)
y_hat_rf_s <- predict(fit.rf, test_dat1, type = "class")
cm_basic_rf <- confusionMatrix(y_hat_rf_s, test_dat1$status) 
```
Plotting: The red curve is the error rate for the Setosa class, the green and blue curves above are for Versicolor and Virginica while the black curve is the Out-of-Bag error rate

## 5.1 Tuning randomForest 

Here, the algorithm is tuned for best accuracy by considering various levels of mtry.

```{r}
#tuning mtry parameter
m <- c(1:11)
fit.rf1 <- sapply(m, function(m){
  fit <- randomForest(status~., data = train_dat_special, mtry = m)
  y_hat <- predict(fit, test_dat1, type="class")
  cm_train <- confusionMatrix(y_hat, test_dat_special$status)
  test_error <- cm_train$overall["Accuracy"]
  tibble(test_error)
  })
  
as.data.frame(fit.rf1)
m[which.max(as.data.frame(fit.rf1))] #accuracy highest when mtry = 7

#checking results
fit.rf.test <- randomForest(status~., data = train_dat_special, mtry = 7)
y_hat.rf <- predict(fit.rf.test, test_dat1, type="class")
cm_test_rf_s <- confusionMatrix(y_hat.rf, test_dat_special$status) 
cm_test_rf_s
cm_test_rf_s$overall["Accuracy"]
F_meas(y_hat.rf, test_dat_special$status) 


#inputting the results
acc_results2 <- bind_rows(acc_results1, 
                          data.frame (Method = c("Selected predictors, RF", 
                                      "Selected predictors, Tuned RF (mtry=7)"), 
                           Accuracy = c(cm_basic_rf$overall["Accuracy"],
                           cm_test_rf_s$overall["Accuracy"]),
                          F1.score  = c(F_meas(y_hat_rf_s,test_dat1$status),
                          F_meas(y_hat.rf,test_dat1$status))))

acc_results2
```
When compared to without mtry tuning, there is much better accuracy and F1 score.  

## 5.2 Further tuning for random forests

Here, various fine tuning for random forest is performed. 

``` {r}
#train control - 5 fold
control <- trainControl(method="cv", number = 5)
grid <- data.frame(mtry = c(1:10))
train_rf <-  train(status~., data = train_dat_special,
                   method = "rf", 
                   ntree = 150,
                   trControl = control,
                   tuneGrid = grid,
                   nSamp = 5000)
ggplot(train_rf)
train_rf

fit.rf2 <- randomForest(status~., data = train_dat_special,
                       minNode = train_rf$bestTune$mtry)
y_hat_rf_s2 <- predict(fit.rf2, test_dat_special)
cm_test_rf2 <- confusionMatrix(y_hat_rf_s2, test_dat_special$status) #acc at 0.8673
cm_test_rf2$overall["Accuracy"]
F_meas(y_hat_rf_s2,test_dat1$status)

#inputting the results
acc_results3 <- bind_rows(acc_results2, 
                          data.frame (Method = "Selected predictors, Tuned RF (5 fold with varying mtry)", 
                           Accuracy = cm_test_rf2$overall["Accuracy"],
                          F1.score  = F_meas(y_hat_rf_s2,test_dat1$status)))

acc_results3


```
# 6. Ensembles

Ensembles are a way to combine different algorithms to produce a better accuracy than a single algorithm. Here, the analysis will combine the top two methods to produce an ensemble. However, this process does not always ensure a better accuracy, as we shall see.

```{r}
#preparing predictions for ensemble comparison
pred <- cbind(y_hat_knn_s, y_hat.rf)
pred <- pred - 1 #not sure why table produces 1 and 2s

#Ensemble Calculation
park <- rowMeans(pred == "1")
y_hat <- ifelse(park > 0.5, "1", "0")
mean(as.factor(y_hat) == test_dat1$status)

#Accuracy at 0.878, lower than tuned randomForest (more methods needed)
#inputting the results
acc_results4 <- bind_rows(acc_results3, 
                          data.frame (Method = "Ensemble of method 2 & 3", 
                           Accuracy = mean(as.factor(y_hat) == test_dat1$status),
                          F1.score  = NA))

acc_results4
```
Above, the ensemble method has a lower accuracy that random forests with tuning.

# **Conclusion**

In conclusion, after comparing two methods in classification, KNN and RF, we see that the RF method with tuning has the highest Accuracy `r cm_test_rf_s$overall["Accuracy"]` and F1 score of `r F_meas(y_hat.rf,test_dat1$status)` for predicting healthy versus Parkinson's patients. However, this accuracy is the same for the tuned KNN and Tuned RF with varying mtry. Yet, this algorithm yields the highest F1 score. 

## **Limitations**
There are two limitations involved in this analysis.

First, while the dataset has many predictors, the number of outcomes is limited (n = 195). Furthermore, the prevalence of status = 1 (parkinsons) is 77% of the overall figure. This could lead to lower F1 scores of sensitivity and specificity, which could affect the efficacy of the model. 

Second, due to the nature of the course, only two main algorithms were used. The caret package offers a wide range of different training options that can be applied. Using these options, together with an ensemble, could yield a higher accuracy for the overall result.

## Future work
Future work could address the above limitations. By working on a larger dataset, reducing the prevalence of one outcome over the other, and including more  algorithms, the accuracy could increase and therefore lead to better detection of parkinson's disease in patients. 

Developing this algorithm with these recommendations would be crucial in diagnosing parkinson's disease, and even work for other related pathologies. A powerful model could lead to early intervention for patients afflicted with these diseases and empower the psychological and medical community better. 

# References

Max A. Little, Patrick E. McSharry, Eric J. Hunter, Lorraine O. Ramig (2008), 'Suitability of dysphonia measurements for telemonitoring of Parkinson's disease', IEEE Transactions on Biomedical Engineering (to appear).

Irizarry, R. (2020) Introduction to Data Science. Retrieved from https://rafalab.github.io/dsbook/ 









